---
title: "Simulation study"
author: "Amanda van Dyk"
date: "27/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This document contains all the code for the simulation study done in Section 3.3 of my research report.

Density of rIG function

```{r}
rig <- function(x,theta, gamma){
  
  inside <- -(x-sqrt(theta*(theta+3*gamma)))^2 #a variable containing the top part of the fraction in the exp
  dens <- sqrt((theta*(theta+3*gamma))/(2*pi*gamma*(x^3)))*exp(inside/(2*gamma*x))
  return(dens)
}
```



Density of Contaminated dist
```{r}
contam <- function(x,theta,gamma,p,beta){
  dens <- p*rig(x,theta,gamma) +(1-p)*rig(x,theta,gamma*beta)
  return(dens)
}
```


Algorithm for contaminated dist:

```{r}
#first have the expectation step
expectationcontam <- function(x,p,theta,gamma,beta){ 
  exqgivenx <- matrix(0,length(x),1) #this will contain all the E[Qk|xk] only 1 col
  
  
  
  for(k in 1:length(x)){
    
    fx <- contam(x[k],theta,gamma,p,beta)
    
    exqgivenx[k] <- (p*rig(x[k],theta,gamma))/fx
    
  }
  
  return(exqgivenx)
  
}



#second we have the maximization step
maximizationcontam <- function(x, exqgivenx, theta_old, gamma_old, beta_old){
  
  p_new <- vector()
  theta_new <- vector()
  gamma_new <-vector()
  #obtaining the new estimates
  #first get the pi values
  m <- length(x)
  
  
  
  p_new <- (1/m)*sum(exqgivenx)
  
  if(p_new<0.5){
    p_new <- 0.5
  }
  
  

    #the log likelihood uses the data given from the maximization     function
    comloglik <- function(par){
  
    theta1 <- par[1]
    gamma1 <- par[2]
    beta1 <- par[3]
    loglik <- 0
    
    if(theta1<=0){
      loglik <- 1000000000
      return(loglik)
    }
    
    if(gamma1<=0){
      loglik <- 100000000
      return(loglik)
    }
    
    if(beta1<=1){
      loglik <- 100000000
      return(loglik)
    }
    
    for(k in 1:m){
       loglik <- loglik + exqgivenx[k]*log(rig(x[k],theta1,gamma1)) +(1-exqgivenx[k])*log(rig(x[k],theta1,gamma1*beta1)) 
      }
    negative <- -loglik
    return(negative)
  }

    
  #I used the pth iteration values for the initial values in optim 
  op <- optim(c(theta_old,gamma_old, beta_old), comloglik,method = "BFGS")
  theta_new <- op$par[1]
  gamma_new <- op$par[2]
  beta_new <-  op$par[3]
  
  
  #the matrix of the parameters
  matrixparm <- cbind(p_new,theta_new,gamma_new, beta_new)
  
  return(matrixparm)
  
  
}


```


Putting the steps together
```{r}
#the incomplete log likelihood used to check convergence:
incomloglikcontam <- function(x,p,theta,gamma,beta){ #g is the number of pdfs in mixture
    m <- length(x)
    loglik <- 0
    sum <-0
    for(k in 1:m){
      
      sum <- p*rig(x[k],theta,gamma) +(1-p)*rig(x[k],theta,gamma*beta) 
      
      loglik <- loglik + log(sum)
    }
    
    return(loglik)
  }




emcontam <- function(x,p_int, theta_int, gamma_int, beta_int){ 
  condition <- 1000
  p <- vector()
  p <- p_int

  
  #first iteration
  exqgivenx <- expectationcontam(x,p,theta = theta_int, gamma = gamma_int, beta=beta_int)
  parameters <- maximizationcontam(x=x,exqgivenx, theta_old = theta_int,gamma_old = gamma_int, beta_old = beta_int)
  
  #the vector containing the loglikelihoods
  loglik1 <- vector()
  loglik_first <- incomloglikcontam(x=x,p =parameters[1,1],theta = parameters[1,2], gamma = parameters[1,3], beta=parameters[1,4])
  i <- 1
  loglik1[i] <- loglik_first
  ai <- vector()
  ai[1] <- 0
  linf <- vector()
  linf[1] <- 0
  parameters_old <- vector()
  
  while(condition>0){
    i <- i+1
    
    parameters_old <- parameters
    #we get the expected values for the next iteration using the prev parameters however not for the expectation func we only give one pi_i value
    exqgivenx <- expectationcontam(x,p=parameters[1,1],theta = parameters[1,2], gamma = parameters[1,3], beta = parameters[1,4])
    #usig prev parameters and expectation func above we get the maximazation
    parameters <- maximizationcontam(x=x,exqgivenx, theta_old = parameters[1,2],gamma_old =parameters[1,3], beta_old = parameters[1,4])
    
    loglik1[i] <- incomloglikcontam(x=x,p=parameters[1,1],theta = parameters[1,2], gamma = parameters[1,3], beta = parameters[1,4])
    
    if(i==3){
      ai[i-1] <- (loglik1[i]-loglik1[i-1])/(loglik1[i-1]-loglik1[i-2])
    }
    
    
    if(i>3){
      ai[i-1] <- (loglik1[i]-loglik1[i-1])/(loglik1[i-1]-loglik1[i-2])
      linf[i-1] <- loglik1[i-2] +(1/(1-ai[i-1]))*(loglik1[i-1]-loglik1[i-2])
      linf[i-2] <- loglik1[i-3] +(1/(1-ai[i-2]))*(loglik1[i-2]-loglik1[i-3])
    }
    
    #print(ai)
    #print(linf)
    #print(loglik1)
    
    if(i>3){
      if(is.na(ai[i-1])){
      condition <- -1
    }
    }
    
    if(i>3){
      if(ai[i-1]==1){
      condition <- -1
    }
    }
    
    if(i>3){
      if(abs(linf[i-1]-linf[i-2])<0.0001){
        condition <- -1
      }
      
    }
    
    

  }
  #print(i)
  #print(loglik1)
  param <- cbind(parameters_old,loglik1[i-1])
  return(param)
}
```

Simulation function:
```{r}
library(VGAM)

simulation <- function(m,theta,gamma, beta, p){ #m is the number of values that need to be simulated
  

  gamma2 <- gamma*beta

  #parameters for ref dist
  mu1 <- sqrt(theta*(theta+3*gamma))
  lambda1 <- (theta*(theta+3*gamma))/gamma

  #parameters for the dist of outliers
  mu2 <- sqrt(theta*(theta+3*gamma2))
  lambda2 <- (theta*(theta+3*gamma2))/gamma2
  
  q1 <- 1-p
  m1 <- round(m*p,0)
  m2 <- round(m*q1,0)
  
  #print(m1)
  #print(m2)
  simdist1 <- rinv.gaussian(n=m1,mu=mu1,lambda=lambda1)
  simdist2 <- rinv.gaussian(n=m2,mu=mu2,lambda=lambda2)

  sim <- c(simdist1, simdist2)

  sim1 <- sample(sim,size=m, replace=FALSE) #so that the values are more random
  #(hist(sim1))
  return(sim1)
}
```


function for initializing the simulation
```{r}

initializecontam2 <- function(sim1){ #sim1 is the simulated values
  # creating a loglikelihood vector to contain the last loglik values
  
  loglik <- vector()
  n <- length(sim1)
  #kmeans initialization
  #print("Initializing using kmeans")
  
  func2 <- matrix(0,2,5)
  
  
#initialization 1:
  #print("Initializing the qk values")
for(k in 1:2){
  n <- length(sim1)

#randomly assign the value 1 or 2 to a specific observation
  q <- sample(c(1,2),size = n, replace=TRUE)
  #print(q)

  pvec <- vector()
  pvec[1] <- length(q[which(q==1)])/n
  pvec[2] <- length(q[which(q==2)])/n
  
  #need the pval to be greater than or equal to 0.5
  if(pvec[1]> pvec[2]){ 
    p <- pvec[1]
    j <- 1 #so that we know which p value is the largest and the other parameters will be associated with them
  }else{
    p<- pvec[2]
    j <- 2}
#the means and var
  means <- vector()
  means[1] <- mean(sim1[which(q==1)])
  means[2] <- mean(sim1[which(q==2)])

  lambdahat <- vector()
  
  if(length(q[which(q==1)])<=1){
    lambdahat[1]<- 0.1 #since beta has to be greater than 1
  }
  else{lambdahat[1] <- (means[1]^3)/(means[1]^3)/var(sim1[which(q==1)])}
  
  if(length(q[which(q==2)])<=1){
    lambdahat[2] <- 0.1
  }
  else{lambdahat[2] <- (means[2]^3)/var(sim1[which(q==2)])}
  
 


#getting the initial gamma and theta values
  theta1 <- vector()
  theta1<- means[j]*(sqrt(1+((9*(means[j]^2))/(4*lambdahat[j]^2)))-(3*means[j])/(2*lambdahat[j])) #use the values associated with the largest p value
  

  gamma1 <- vector()
  gamma1[1] <- (means[1]^2)/lambdahat[1]
  gamma1[2] <- (means[2]^2)/lambdahat[2]
  
  

  
  if(gamma1[1]>gamma1[2]){
    beta1 <- gamma1[1]/gamma1[2]
    gamma11 <- gamma1[2]
  }else{beta1 <- gamma1[2]/gamma1[1]
    gamma11 <- gamma1[1]}
  
  
  func2[k,] <- emcontam(x=sim1,p_int=p, theta_int=theta1, gamma_int=gamma11, beta_int = beta1)
  #print("Initial :")
  #print(g)
  #print(func2)
  loglik[k] <- func2[k,5]
  }
  highestloklik <- max(loglik)
  highestinitial <- which.max(loglik)
  #rbind(func,func2)
  return(func2[highestinitial,])
}
```





Simulating for different p values
```{r}
set.seed(500) # so that the results can be replicated

Nit <- 200 #the number of iterations

p1 <- 0.7
p2 <- 0.8
p3 <- 0.9

parameters21 <- matrix(0,Nit,5)
parameters22 <- matrix(0,Nit,5)
parameters23 <- matrix(0,Nit,5)

n <- 500
theta <- 1
gamma <- 1
beta <- 50
 

for(i in 1:Nit){
  print(i)
  sim <- simulation(n,theta,gamma,beta,p1)
  parameters21[i,] <- initializecontam2(sim)
}

for(i in 1:Nit){
  print(i)
  sim <- simulation(n,theta,gamma,beta,p2)
  parameters22[i,] <- initializecontam2(sim)
}

for(i in 1:Nit){
  print(i)
  sim <- simulation(n,theta,gamma,beta,p3)
  parameters23[i,] <- initializecontam2(sim)
}
```


Simulating for different betas
```{r}
set.seed(500) # so that the results can be replicated

Nit <- 200 #the number of iterations
beta1 <- 20
beta3 <- 70

parameters31 <- matrix(0,Nit,5) #saving the different parameter values in here
parameters32 <- matrix(0,Nit,5)


n <- 500
theta <- 1
gamma <- 1
p <- 0.8 

for(i in 1:Nit){
  print(i)
  sim <- simulation(n,theta,gamma,beta1,p)
  parameters31[i,] <- initializecontam2(sim)
}

for(i in 1:Nit){
  print(i)
  sim <- simulation(n,theta,gamma,beta3,p)
  parameters32[i,] <- initializecontam2(sim)
}


```


Obtaining the averages of the estimated parameters and the standard deviations:

```{r}
#creating an extra column for gamma*beta
parameters21new <- cbind(parameters21, parameters21[,3]*parameters21[,4])

parameters22new<- cbind(parameters22, parameters22[,3]*parameters22[,4])

parameters23 <- cbind(parameters23, parameters23[,3]*parameters23[,4])

parameters31 <- cbind(parameters31, parameters31[,3]*parameters31[,4])

parameters32 <- cbind(parameters32, parameters32[,3]*parameters32[,4])

```

The averages and standard deviations for the different p values:
```{r}
#for the diff p-vals:
means1 <- colMeans(parameters21new)
means2 <- colMeans(parameters22new)
means3 <- colMeans(parameters23)



means1
means2
means3


standarddev1 <- matrix(0,1,5)
standarddev1[1,1] <- sqrt(var(parameters21new[,1]))
standarddev1[1,2] <- sqrt(var(parameters21new[,2]))
standarddev1[1,3] <- sqrt(var(parameters21new[,3]))
standarddev1[1,4] <- sqrt(var(parameters21new[,4]))
standarddev1[1,5] <- sqrt(var(parameters21new[,6]))

standarddev1

standarddev2 <- matrix(0,1,5)
standarddev2[1,1] <- sqrt(var(parameters22new[,1]))
standarddev2[1,2] <- sqrt(var(parameters22new[,2]))
standarddev2[1,3] <- sqrt(var(parameters22new[,3]))
standarddev2[1,4] <- sqrt(var(parameters22new[,4]))
standarddev2[1,5] <- sqrt(var(parameters22new[,6]))

standarddev2

standarddev3 <- matrix(0,1,5)

standarddev3[1,1] <-sqrt(var(parameters23[,1]))
standarddev3[1,2] <- sqrt(var(parameters23[,2]))
standarddev3[1,3] <- sqrt(var(parameters23[,3]))
standarddev3[1,4] <-sqrt(var(parameters23[,4]))
standarddev3[1,5] <- sqrt(var(parameters23[,6]))

standarddev3
```

The averages and standard deviations for the different beta values:
```{r}
#for diff beta values
means4 <- colMeans(parameters31)
means5 <- colMeans(parameters32)


means4
means5


standarddev4 <- matrix(0,1,5)
standarddev4[1,1] <- sqrt(var(parameters31[,1]))
standarddev4[1,2] <- sqrt(var(parameters31[,2]))
standarddev4[1,3] <- sqrt(var(parameters31[,3]))
standarddev4[1,4] <- sqrt(var(parameters31[,4]))
standarddev4[1,5] <- sqrt(var(parameters31[,6]))

standarddev4

standarddev5 <- matrix(0,1,5)
standarddev5[1,1] <- sqrt(var(parameters32[,1]))
standarddev5[1,2] <- sqrt(var(parameters32[,2]))
standarddev5[1,3] <- sqrt(var(parameters32[,3]))
standarddev5[1,4] <- sqrt(var(parameters32[,4]))
standarddev5[1,5] <- sqrt(var(parameters32[,6]))

standarddev5
```
